---
title: "Introduction to didTEnets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to didTEnets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Introduction to didTEnets

The `didTEnets` package implements the **Transfer Entropy-informed Difference-in-Differences (TE-DiD)** framework for policy evaluation in interconnected systems. This methodology combines network analysis with causal inference to measure policy effects on financial contagion and systemic risk transmission.

## Overview

Traditional difference-in-differences (DiD) approaches often use simple outcome variables like stock returns or volatility, which fail to capture the complex dynamics of systemic risk transmission in interconnected financial markets. The TE-DiD framework addresses this limitation by:

1. **Network-Informed Outcomes**: Using the Network-Informed Contagion Index (NICI) instead of simple returns
2. **Directed Information Flow**: Measuring transfer entropy to quantify directional risk transmission
3. **Policy Spillovers**: Accounting for spillover effects through the Network-Informed Policy Intensity (NIPI)
4. **Crisis Analysis**: Focusing on extreme events through quantile conditioning

## Key Concepts

### Transfer Entropy (TE)
Transfer entropy measures the directed flow of information between time series, quantifying how much knowing the past of series X reduces uncertainty about the future of series Y, beyond what we already know from Y's own past.

### Network-Informed Contagion Index (NICI)
NICI serves as the outcome variable in TE-DiD models. It measures the weighted sum of incoming transfer entropy flows to a country, representing how much systemic risk the country is absorbing from its network neighbors.

### Network-Informed Policy Intensity (NIPI)
NIPI serves as the treatment variable, combining direct policy effects with spillover effects from neighboring countries in the network.

## Installation

You can install the development version of didTEnets from GitHub:

```{r eval=FALSE}
# Install devtools if you haven't already
install.packages("devtools")

# Install didTEnets
devtools::install_github("avishekb9/didTEnets")
```

```{r setup}
library(didTEnets)
library(ggplot2)
library(dplyr)
```

## Quick Start Example

Here's a basic example using the package's sample data:

```{r eval=FALSE}
# Load sample data
data(sample_financial_returns)
data(sample_policy_data)

# Quick analysis with built-in sample data
tickers <- c("^GSPC" = "USA", "^FTSE" = "GBR", "^GDAXI" = "DEU", 
             "^FCHI" = "FRA", "^N225" = "JPN", "^AXJO" = "AUS")

# Run complete TE-DiD analysis
results <- run_complete_tedid_analysis(
  tickers = tickers,
  start_date = "2019-01-01",
  end_date = "2021-12-31",
  create_plots = TRUE,
  verbose = TRUE
)

# View main results
summary(results$tedid$main_model)
```

## Step-by-Step Workflow

### 1. Data Acquisition

```{r eval=FALSE}
# Define financial market tickers
tickers <- c("^GSPC" = "USA", "^FTSE" = "GBR", "^GDAXI" = "DEU")

# Download financial data
financial_data <- download_financial_data(tickers, "2019-01-01", "2021-12-31")
```

### 2. Policy Data Creation

```{r eval=FALSE}
# Create realistic policy data
policy_data <- create_policy_data(financial_data$countries, "2019-01-01", "2021-12-31")
```

### 3. Network Weights Calculation

```{r eval=FALSE}
# Calculate different types of network weights
network_weights <- calculate_network_weights(financial_data$returns)
```

### 4. Transfer Entropy Analysis

```{r eval=FALSE}
# Define periods for analysis
crisis_start <- as.Date("2020-03-01")
crisis_end <- as.Date("2020-12-31")

period_breaks <- list(
  "Pre_Crisis" = which(index(financial_data$returns) < crisis_start),
  "Crisis" = which(index(financial_data$returns) >= crisis_start & 
                  index(financial_data$returns) <= crisis_end),
  "Post_Crisis" = which(index(financial_data$returns) > crisis_end)
)

# Calculate TE matrices
te_matrices <- calculate_te_matrices_by_period(financial_data$returns, period_breaks)
```

### 5. NICI and NIPI Construction

```{r eval=FALSE}
# Calculate outcome variable (NICI)
nici_data <- calculate_nici(te_matrices, network_weights)

# Prepare monthly policy data
monthly_policy <- policy_data %>%
  mutate(year_month = format(Date, "%Y-%m")) %>%
  group_by(CountryCode, year_month) %>%
  summarise(StringencyIndex = mean(StringencyIndex, na.rm = TRUE), .groups = "drop")

# Calculate treatment variable (NIPI)
nipi_data <- calculate_nipi(monthly_policy, network_weights)
```

### 6. TE-DiD Model Estimation

```{r eval=FALSE}
# Prepare dataset for TE-DiD estimation
tedid_dataset <- prepare_tedid_dataset(nici_data, nipi_data, treatment_groups)

# Estimate TE-DiD model
tedid_model <- estimate_tedid_model(tedid_dataset)

# Extract and interpret results
results <- extract_tedid_results(tedid_model, tedid_dataset)
```

### 7. Visualization

```{r eval=FALSE}
# Create network comparison plots
network_plots <- create_network_comparison(te_matrices)

# Create TE-DiD result visualizations
tedid_plots <- visualize_tedid_results(tedid_dataset)

# Display plots
print(tedid_plots$timeseries)
print(tedid_plots$difference)
```

## Interpreting Results

The key output of a TE-DiD analysis is the interaction coefficient β₁ from the model:

**IncomingFlow_{it} = β₀ + β₁ · EarlyAdopter_i × CrisisPeriod_t + α_i + γ_t + ε_{it}**

Where:
- **β₁ < 0**: Early policy adoption **reduced** incoming financial contagion
- **β₁ > 0**: Early policy adoption **increased** incoming financial contagion (possibly due to uncertainty)
- **β₁ ≈ 0**: No significant policy effect

## Advantages over Traditional DiD

1. **Network-Aware**: Captures interconnectedness and spillover effects
2. **Crisis-Focused**: Uses quantile conditioning to focus on extreme events
3. **Directional**: Measures directed information flows rather than simple correlations
4. **Policy-Relevant**: Provides clear implications for crisis management

## Next Steps

- See the **Network Analysis** vignette for detailed network construction methods
- See the **Policy Evaluation** vignette for advanced TE-DiD specifications
- Check the **Case Studies** vignette for real-world applications

## References

1. Schreiber, T. (2000). Measuring information transfer. *Physical Review Letters*, 85(2), 461.
2. Angrist, J. D., & Pischke, J. S. (2009). *Mostly harmless econometrics: An empiricist's companion*. Princeton University Press.
3. Hale, T., Angrist, N., Goldszmidt, R., et al. (2021). A global panel database of pandemic policies (Oxford COVID-19 Government Response Tracker). *Nature Human Behaviour*, 5, 529-538.