---
title: "Network Analysis with didTEnets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Network Analysis with didTEnets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/",
  out.width = "100%"
)
```

# Network Analysis with didTEnets

This vignette demonstrates the network analysis capabilities of the `didTEnets` package, focusing on transfer entropy calculation, network weight construction, and network evolution visualization.

```{r setup}
library(didTEnets)
library(ggplot2)
library(dplyr)
library(igraph)
library(ggraph)
```

## Understanding Transfer Entropy

Transfer entropy (TE) is a model-free measure of directed information flow between time series. Unlike correlation, which is symmetric, transfer entropy captures the direction of information flow.

### Basic Transfer Entropy Calculation

```{r eval=FALSE}
# Generate sample data with known relationship
set.seed(123)
n <- 200
source <- rnorm(n)
target <- 0.3 * source + rnorm(n, 0, 0.8)  # target influenced by source

# Calculate transfer entropy in both directions
te_source_to_target <- calculate_enhanced_transfer_entropy(source, target)
te_target_to_source <- calculate_enhanced_transfer_entropy(target, source)

cat("TE from source to target:", round(te_source_to_target, 4), "\n")
cat("TE from target to source:", round(te_target_to_source, 4), "\n")
```

The enhanced transfer entropy function in `didTEnets` includes several innovations:

1. **Quantile Conditioning**: Focus on extreme events (crisis periods)
2. **Volatility Scaling**: Incorporate market volatility for realistic magnitudes
3. **Robust Handling**: Deal with missing data and insufficient observations

### Quantile Conditioning for Crisis Analysis

```{r eval=FALSE}
# Example with crisis periods
set.seed(456)
n <- 300
normal_period <- rnorm(200, 0, 1)
crisis_period <- rnorm(100, -2, 2)  # More volatile, negative bias
source_series <- c(normal_period, crisis_period)
target_series <- c(rnorm(200, 0, 1), 0.6 * crisis_period + rnorm(100, 0, 0.5))

# Calculate TE focusing on crisis events (bottom 10% quantile)
te_crisis <- calculate_enhanced_transfer_entropy(source_series, target_series, 
                                               quantile_threshold = 0.1)

# Calculate TE focusing on normal periods (50th percentile)
te_normal <- calculate_enhanced_transfer_entropy(source_series, target_series, 
                                               quantile_threshold = 0.5)

cat("TE during crisis events:", round(te_crisis, 4), "\n")
cat("TE during normal periods:", round(te_normal, 4), "\n")
```

## Network Weight Construction

Network weights determine how policy spillovers and contagion effects propagate through the financial system. The `didTEnets` package supports multiple weight types:

### 1. Correlation-Based Weights

```{r eval=FALSE}
# Using sample data
data(sample_financial_returns)

# Calculate correlation-based weights
cor_weights <- calculate_network_weights(sample_financial_returns, "correlation")

# Examine the weight matrix
print(round(cor_weights$correlation, 3))
```

Correlation-based weights capture co-movement patterns between markets. Higher correlations indicate stronger connections.

### 2. Geographic Weights

```{r eval=FALSE}
# Calculate geographic weights with custom regions
custom_regions <- list(
  europe = c("GBR", "DEU", "FRA"),
  asia_pacific = c("JPN", "AUS"),
  americas = c("USA")
)

geo_weights <- calculate_network_weights(sample_financial_returns, "geographic",
                                       geographic_groups = custom_regions)

print(round(geo_weights$geographic, 3))
```

Geographic weights assume that countries in the same region have stronger connections due to trade, cultural, and institutional similarities.

### 3. Economic Similarity Weights

```{r eval=FALSE}
# Calculate economic similarity weights (based on volatility patterns)
econ_weights <- calculate_network_weights(sample_financial_returns, "economic")

print(round(econ_weights$economic, 3))
```

Economic similarity weights use market volatility patterns as a proxy for economic development level and market maturity.

### 4. Composite Weights

```{r eval=FALSE}
# Calculate all weight types and create composite
all_weights <- calculate_network_weights(sample_financial_returns)

# The composite weights combine multiple sources of information
print(round(all_weights$composite, 3))
```

## Transfer Entropy Matrix Calculation

For network analysis, we calculate transfer entropy between all pairs of countries for different time periods and market conditions:

```{r eval=FALSE}
# Define analysis periods
returns_data <- sample_financial_returns
n_obs <- nrow(returns_data)

period_breaks <- list(
  "Pre_Crisis" = 1:floor(n_obs * 0.4),
  "Crisis" = floor(n_obs * 0.4):floor(n_obs * 0.7),
  "Post_Crisis" = floor(n_obs * 0.7):n_obs
)

# Calculate TE matrices for different periods and quantiles
te_matrices <- calculate_te_matrices_by_period(returns_data, period_breaks, 
                                             quantiles = c(0.1, 0.5, 0.9))

# Examine the structure
names(te_matrices)
```

Each matrix shows directed information flows between countries:
- **Rows**: Source countries (information senders)
- **Columns**: Target countries (information receivers)
- **Values**: Transfer entropy strength

## Network Statistics and Evolution

### Basic Network Statistics

```{r eval=FALSE}
# Calculate comprehensive network statistics
network_stats <- calculate_network_statistics(te_matrices)

# View summary statistics
print(network_stats)
```

Key network statistics include:
- **Density**: Proportion of possible connections that exist
- **Transitivity**: Clustering coefficient (A→B→C implies A→C)
- **Average Strength**: Mean TE flow strength
- **Max TE**: Strongest individual connection

### Contagion Analysis

```{r eval=FALSE}
# Analyze network evolution to establish contagion
crisis_stats <- network_stats %>%
  filter(Quantile == "Q0.1") %>%  # Focus on crisis quantile
  arrange(Period)

if (nrow(crisis_stats) >= 2) {
  pre_crisis <- crisis_stats[crisis_stats$Period == "Pre_Crisis", ]
  crisis <- crisis_stats[crisis_stats$Period == "Crisis", ]
  
  if (nrow(pre_crisis) > 0 && nrow(crisis) > 0) {
    density_change <- crisis$Density - pre_crisis$Density
    mean_te_change <- crisis$Mean_TE - pre_crisis$Mean_TE
    
    cat("Network density change:", round(density_change, 4), "\n")
    cat("Mean TE change:", round(mean_te_change, 4), "\n")
    
    if (density_change > 0.05 || mean_te_change > 0.01) {
      cat("✓ CONTAGION ESTABLISHED\n")
    } else {
      cat("○ Limited evidence of contagion\n")
    }
  }
}
```

## Network Visualization

### Network Evolution Plots

```{r eval=FALSE}
# Create network comparison visualizations
network_plots <- create_network_comparison(te_matrices, threshold = 0.02)

# Display evolution across periods
if (length(network_plots) > 0) {
  print(network_plots$Pre_Crisis)
  print(network_plots$Crisis)
  print(network_plots$Post_Crisis)
}
```

### Transfer Entropy Heatmaps

```{r eval=FALSE}
# Create TE heatmaps for different periods
te_heatmaps <- create_te_heatmaps(te_matrices)

# Display heatmaps
if (length(te_heatmaps) > 0) {
  print(te_heatmaps$Pre_Crisis)
  print(te_heatmaps$Crisis)
}
```

### Custom Network Visualization

```{r eval=FALSE}
# Create custom network plot
create_custom_network_plot <- function(te_matrix, title = "Financial Network") {
  # Convert to igraph object
  g <- graph_from_adjacency_matrix(te_matrix, mode = "directed", 
                                  weighted = TRUE, diag = FALSE)
  
  # Set node attributes
  V(g)$size <- degree(g, mode = "in") * 5 + 10
  V(g)$color <- case_when(
    V(g)$name %in% c("USA", "GBR") ~ "red",
    V(g)$name %in% c("DEU", "FRA") ~ "blue",
    TRUE ~ "orange"
  )
  
  # Create plot
  ggraph(g, layout = "stress") +
    geom_edge_arc(aes(width = weight, alpha = weight),
                 arrow = arrow(length = unit(2, "mm")),
                 color = "gray50") +
    geom_node_point(aes(size = size, color = color), alpha = 0.8) +
    geom_node_text(aes(label = name), repel = TRUE, fontface = "bold") +
    scale_edge_width_continuous(range = c(0.5, 2), guide = "none") +
    scale_edge_alpha_continuous(range = c(0.3, 0.8), guide = "none") +
    scale_size_continuous(range = c(5, 15), guide = "none") +
    scale_color_identity() +
    labs(title = title) +
    theme_void()
}

# Apply to crisis period matrix
if ("Crisis_Q0.1" %in% names(te_matrices)) {
  crisis_plot <- create_custom_network_plot(te_matrices$Crisis_Q0.1, 
                                           "Crisis Period Network")
  print(crisis_plot)
}
```

## Advanced Network Analysis

### Centrality Measures

```{r eval=FALSE}
# Calculate network centrality measures
calculate_centrality_measures <- function(te_matrix) {
  g <- graph_from_adjacency_matrix(te_matrix, mode = "directed", 
                                  weighted = TRUE, diag = FALSE)
  
  centrality_df <- data.frame(
    Country = V(g)$name,
    InDegree = degree(g, mode = "in"),
    OutDegree = degree(g, mode = "out"),
    InStrength = strength(g, mode = "in"),
    OutStrength = strength(g, mode = "out"),
    Betweenness = betweenness(g),
    PageRank = page_rank(g)$vector
  )
  
  return(centrality_df)
}

# Apply to crisis matrix
if ("Crisis_Q0.1" %in% names(te_matrices)) {
  centrality_measures <- calculate_centrality_measures(te_matrices$Crisis_Q0.1)
  print(centrality_measures)
}
```

### Network Persistence Analysis

```{r eval=FALSE}
# Analyze which connections persist across periods
analyze_network_persistence <- function(te_matrices, threshold = 0.05) {
  # Get crisis period matrices
  crisis_matrices <- te_matrices[grep("Q0.1", names(te_matrices))]
  
  # Binarize matrices based on threshold
  binary_matrices <- lapply(crisis_matrices, function(x) {
    (x > threshold) * 1
  })
  
  # Calculate persistence (connections present in all periods)
  if (length(binary_matrices) >= 2) {
    persistence_matrix <- Reduce("+", binary_matrices) / length(binary_matrices)
    
    # Find highly persistent connections (present in >75% of periods)
    persistent_connections <- which(persistence_matrix > 0.75, arr.ind = TRUE)
    
    if (nrow(persistent_connections) > 0) {
      cat("Highly persistent connections:\n")
      for (i in 1:nrow(persistent_connections)) {
        from <- rownames(persistence_matrix)[persistent_connections[i, 1]]
        to <- colnames(persistence_matrix)[persistent_connections[i, 2]]
        cat(paste(from, "→", to), "\n")
      }
    }
  }
}

# Analyze persistence
analyze_network_persistence(te_matrices)
```

## Best Practices

### 1. Data Quality
- Ensure sufficient observations for reliable TE estimation (minimum 30-50 per period)
- Handle missing data appropriately
- Check for structural breaks in the data

### 2. Parameter Selection
- Choose quantile thresholds based on research question (0.1 for crisis, 0.5 for normal)
- Set appropriate TE thresholds for network visualization (typically 0.01-0.05)
- Consider lag structure for TE calculation

### 3. Interpretation
- Remember that TE measures information flow, not causation
- Consider economic mechanisms behind network connections
- Validate results with known financial relationships

### 4. Computational Considerations
- TE calculation can be computationally intensive for large networks
- Consider parallel processing for multiple periods/quantiles
- Cache results for repeated analysis

## Conclusion

The network analysis capabilities of `didTEnets` provide powerful tools for understanding financial contagion and policy transmission mechanisms. By combining transfer entropy with network analysis, researchers can:

1. Identify key transmission channels during crisis periods
2. Measure the evolution of financial networks over time
3. Quantify the importance of different countries as sources or receivers of risk
4. Visualize complex network structures in an intuitive way

These insights are crucial for designing effective policy interventions and understanding systemic risk in interconnected financial systems.